---
title: "MakeSearchSim Package Overview"
author: "Sarah C. Jenson"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: TRUE
vignette: >
  %\VignetteIndexEntry{MakeSearchSim Package Overview}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  comment = "#"
)
library(stringi)
library(MakeSearchSim)
```

# Overview
MakeSearchSim contains various functions and workflows which have the end goal
of modeling the de novo results of an MS/MS dataset representing multiple
organisms. This is accomplished by combining existing MS/MS datasets either on
the level of the mzML files or by mixing pre-existing de novo results. One MS/MS
dataset is designated as the "primary" dataset and all scans from that dataset
are included in the simulation. A second MS/MS dataset is designated as the
"contributor" dataset and only a portion of the scans from this dataset are used
in the simulation. Scans from the contributor dataset are randomly selected
using MSGF+ results to correspond to a user specified number of proteins and
number of peptides per protein.

When combining at the level of the mzML file, a new mzML file is created by
appending the subset of scans from the contributor dataset to the primary
dataset using MSConvert. 

The resulting simulated mzML file is then analyzed
using Novor to create the de novo results. This method is relatively slow
because of the need to create and analyze a new mzML file for each simulation.
Combining the primary and contributor datasets by mixing pre-existing de novo
results is much faster as it is essentially just subsetting and merging
dataframes.


This vignette focuses on the most basic functions contained in MakeSearchSim.
How these functions are connected to make the main high throughput workflows is
described in separate vignettes you can access using the below command.
```{r eval=FALSE}
browseVignettes("MakeSearchSim")
```


#### Note on terms

**Dataset vs Filename:**  
Throughout the documentation and code I use the term "Dataset" or "base name" to
refer to the name of a file *without* the extension. If I'm referring to the
file *with* extension, I'll use the term "Filename".

This is because I often have to convert the same raw data into different formats
with different extensions. I also use the same base dataset name with various
suffixes to denote what kind of results the file contains. For example, the
Novor results for the dataset "Alpha_Beta_Sim42" would be called
"Alpha_Beta_Sim42_Novor.csv".

**SimID:**  
The SimID is a number used as a unique identifier for the combination of
parameters which were used to make a simulated dataset. Even though the name of
the simulated dataset contains some information about the datasets contents,
**under no circumstances should you use the same SimID twice**. This ensures
that wherever the simulated file (or it's associated results) ends up, there
will never be any ambiguity on what is in the simulated dataset. The SimID is
also used as the seed for the sample() function when selecting contributor
scans.


# Selecting contributor scans
Selecting scans to use from a contributor dataset is done using `selectScans()`.
This function requires a dataframe of "unrolled"" MSGF+ results, meaning each
row must represent a single unique combination of scan, peptide, and protein.
This differs from "rolled" results, where multiple proteins are referenced in a
single row. The input dataframe must have the following columns (column names
must exactly match):


* cleanseq: contains the peptide sequence without modification info
* Protein: contains the accession or some unique identifier for the protein
* ScanNum: contains the scan numbers  


`selectScans()` returns a set of scan numbers associated with a randomly
selected set of proteins, the number of which is set by the user using the Nprot
parameter. Only proteins with at least 2 peptides detected in the MSGF+ results
are considered.

For each randomly selected protein, the user can specify how many peptides per
protein are considered using the `Mpep` parameter. For example, if a protein
with 10 peptides is selected, `Mpep = 5` would randomly select 5 of those
peptides. All scans associated with those 5 peptides would then be added to the
list of scans to be returned. If a protein has fewer peptides than `Mpep`, all
peptides from that protein are used and all scans associated with each peptide
are returned.


#### Making scan selection reproducable across R versions
The output of `sample()` depends on two external factors: an integer value
called the seed, and the RNG (Random Number Generator) state, which a **global
variable** that defines what algorithms and parameters are used for random
number generation. `sample()` will always give the same results for given set of
inputs if the seed and RNG state are held constant. These factors are set by
separate functions called before `sample()`.

`selectScans()` uses the SimID as the seed as it is already recorded througout
the pipeline. By default the seed is based on the time and date `sample()` was
run, so without manually setting the seed the output of `selectScans()` would
not be reproducable.

**The default RNG state was changed in R 3.6.0**, which means that even with the
same seed, the outputs of `sample()` will not match those run with previous
version of R unless the RNG state is manually set. To make the output of
`selectScans()` reproducable across R versions, the RNG state is set to
corrispond to the version of R contained in the `rng_version` parameter. **The
RNG state is a global variable**, so to avoid side efects the RNG state is
restored its previous value once `selectScans()` exits. See `?RNGkind` for more
details about the RNG state.




Below is an example call:
```{r eval=FALSE}
selectScans(
  databaseResults = MSGF1,
  Nprot = 20,
  Mpep = 5,
  SimID = 42,
  rng_version = "R_version_default"
)
```

# Creating simulated mzMLs
`SubsetMergeMzmls()` uses the filter and merge commands of MSConvert to create a
new simulated mzML file from a primary and contributor mzML file.
`SubsetMergeMzmls()` first makes a new mzML containing the specified set of
scans from the contributor mzML using the MSConvert `filter` parameter. This
mzML is named using the `simID` parameter like so: [Contributor
dataset]_SimID[SimID]_subset.mzML.
It then merges this mzML with the primary mzML to make a new simulated mzML
which is named using the `simname` parameter. The primary and contributor mzMLs
must be in the working directory. MSConvert must be installed and in the
computer's PATH.

To set the computer's PATH, see
https://stackoverflow.com/questions/43882307/where-should-i-set-the-variable-path-in-r for instructions.

`SubsetMergeMzmls()` is called like so:
```{r eval=FALSE}
SubsetMergeMzmls(
  scans = scanvector,
  contributorName = "ContributorDataset",
  primaryName = "PrimaryDataset",
  simname = "SimulatedFile.mzML",
  simID = 42
)
```


# Mixing exsisting de novo results
`MixNovor()` returns a dataframe of simulated Novor results by mixing Novor
results from the primary and contributor datasets. Novor results are passed to
the function as a list of dataframes (`NovorResultList`), which may contain
Novor results for many different datasets. The function then uses the dataset
names in `PrimaryDataset` and `ContributorDataset` to extract the correct
dataframe from the list. It does this like so:

```
contributorNovor<-NovorResultList[[match(ContributorDataset,
                                    names(NovorResultList))]]`
```

Which means the elements in the NovorResultList must be named according to their
dataset (see the Note on terms above). There are checks within the function to
assure this is the case.

Next, `MixNovor()` subsets the contributor Novor results using the scan numbers
in `contributorScans`and appends them to the Novor results of the primary
dataset. **Note:** To avoid duplication of scan numbers, it is recommended to
have a column in the input Novor result dataframes which lists the original
file. This is added if you use `importSingleNovor()` to import the Novor result
from the csv files.

`MixNovor()` is called like so:
```{r eval=FALSE}
MixNovor(
  PrimaryDataset = "PrimaryDataset",
  ContributorDataset = "ContributorDataset",
  NovorResultList = ListOfNovorDataframes,
  SimID = 42
)
```

# Importing raw Novor Results
ImportSingleNovor() does the following things: 

  + reads raw csv  
  + cleans up column names  
  + adds peptide sequence with no modifications (cleanseq)  
  + adds column for dataset name of the novor result file (NovorDataset)   
  + calls `SingleMGFIDtoScan()` to add scan numbers and source file names  
    

The "id" column in the raw Novor results represents the position of the scan in
the MGF file. The "scanNum" column in the raw csv will be all 0's because the
MGF file created by MSConvert does not contain a `scan= ` line which Novor needs
to extract scan numbers. 
So the `SingleMGFIDtoScan()`function creates a table
which correlates the position of the scan in the MGF file to the scan number and
original file name contained in each title line. It expects the scan titles to
be in the form made by MSConvert using this filter: 

```
filter="titleMaker <RunId>.<ScanNumber>.<ScanNumber>.<ChargeState> File:"<SourcePath>", NativeID:"<Id>"` 
```

These scan titles also contain the name of the original source
file, so we can differentiate between scans from the contributor and primary
datasets in Novor results from a simulated mzML.

# Digesting peptides
Some analysis methods will only accept fully tryptic peptides with no missed
cleavages. `digestPeptides()` preforms an *in silico* tryptic digest on each
peptide by splitting the peptide after R or K (but not RP or KP if
`prolineBlock=TRUE`). Any resultant peptides which have a length less than the
`discardLessThan` parameter are discarded.

```{r}
rawpeps <-
  c("IVGRNGLCVDVR",
    "NGLCVDVRDGR",
    "FHNGNAIQLWPCKSNTDANQLWTLK",
    "VGLPINQR")
digestPeptides(rawpeps)
```





