---
title: "Make simulations from many organisms using Novor results"
author: "Sarah C. Jenson"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: TRUE
vignette: >
  %\VignetteIndexEntry{Make simulations from many organisms using Novor results}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
library(plyr)
library(reshape2)
library(assertthat)
library(knitr)
library(MakeSearchSim)

knitr::opts_chunk$set(
  comment = "#"
)
```

# Overview
This vignette describes a workflow for making simulated peptide lists by mixing
Novor results from an arbitrary number of primary and contributor datasets.
Other workflows are limited to one primary dataset (from which all scans are
included) and one contributor dataset (from which a portion of scans are
included).

This vignette covers the use of the various functions which build on the basic
functions described in [MakeSearchSim Package
Overview](makesearchsim-overview.html) to make them part of a high throughput
pipeline. It is recommended you read that vignette first.

An Rmarkdown template to run this workflow called "Multiple Organism Simulations
using Novor Results" can be accessed using File -> New File -> R Markdown ->
From Template.


# Inputs
1. **A csv file(SimRecipes)** which contains at least the following columns.
Each section of rows with the same SimID corrisponds to one simulation.
    + SimID: Numeric identifier unique to each simulation  
    + Dataset: Name of a dataset, has no extension or "_Novor" suffix 
    + Type: Either "Primary" or "Contributor" 
    + Organism: name of the organism associated with the dataset
    + TargetNumContributorProteins: How many proteins to add from the contributor dataset  (This should be "NA" for Primary datasets)
    + TargetContributorPep_per_Prot: How many peptides per protein to add from each selected contributor protein (This should be "NA" for Primary datasets) 

2. **MSGF+ results** as a list of dataframes or a single dataframe with a
dataset column that can be converted to a list. Each entry in the list contains
the MSGF+ results for a contributor dataset. The name of the dataframe in the
list must match to the entry in the Dataset column of SimRecipes. The dataframe
must have the following columns:
    + Dataset: contains the dataset name
    + cleanseq: contains a clean peptide sequence (no modification info)
    + ScanNum: contains scan numbers 
    + Protein: contains protein accessions 
    + **Note:** Data must be unrolled, meaning one row for each scan, peptide, and protein combination present in the data
    
3. **Novor results:** This vignette assumes you already have Novor results for
each dataset in the appropriate format. See the [Simulation workflow using
existing Novor results](simulate-via-existing-novor.html) vignette for a
description of how to run Novor and/or import the results.





### Loading inputs
An example SimRecipes table is included with the package. 
```{r}
#loading SimRecipes included with package
SimRecipes <- MakeSearchSim::SimRecipes5_8_long
knitr::kable(SimRecipes)
```

The MSGF+ and raw Novor results are very large and are not included with the
package.
```{r eval=FALSE}
#loading Novor Results
load("SimID_5_8_RawNovorList.RData")

#loading MSGF+ results
load("dbResultList_Sim5_8.RData")
```

# Selecting Scans
Call `SingleDataset_SelectScans` on each row of the SimRecipes table using
`adply` and passing the MSGF+ result list using the `dbResultList` parameter.
See:
```{r eval=FALSE}
multidataTracker <- adply(
  SimRecipes5_8_long,
  1,
  SingleDataset_SelectScans,
  dbResultList = dbResultList_Sim5_8,
  rng_version = "R_version_default"
)
```

The output for each row is combined into a new dataframe by `adply`. This
dataframe contains what scans were selected from the contributor datasets.
Primary datasets contain NA in these columns because the simulation will use all
scans from these datasets.

Six new columns are added
```{r}
colnames(MakeSearchSim::multiDataTracker_SimID5_8)[8:13]
```

You can inspect the whole dataTracker dataframe like so:

```{r eval=FALSE}
View(MakeSearchSim::multiDataTracker_SimID5_8)
```


# Mixing Novor Results
The next step is to combine Novor PSMs from each dataset using the information
stored in the dataTracker dataframe created above. All Novor PSMs are used from
datasets marked as "Primary" in the Type column. For "Contributor" datasets,
only PSMs associated with scans in the "ContributorScans" column are used in the
simulation.

To do this, call  `MixMultipleNovor` on the `multidataTracker` dataframe created
above using `dlply` and splitting by the SimID. `dlply` will split
`multidataTracker` by SimID and call `MixMultipleNovor` on each chunck. The
results from each call are combined into a list with each entry corrisponding to
a different simulation.
```{r eval=FALSE}
RawNovorMix_5_8 <- dlply(multidataTracker,
                         .(SimID),
                         MixMultipleNovor,
                         RawNovorList = RawNovorList)
```

# Filtering Novor Results
`filterNovor` filters raw Novor results by user specified cutoffs of ppm error,
Novor Score, and peptide length. Peptide length is calculated within the
function based on the "cleanseq" column and added as a "length" column.

The raw Novor results must have columns called ""cleanseq", "NovorScore", and
"ppmError".

Here `llply` calls `filterNovor` on each entry in `RawNovorMix_5_8` and combines
the results into a list. Each entry in the list is a dataframe representing all
the filtered PSMs for that simulation. Each entry is named after the SimID.
```{r eval=FALSE}
filtNovorMix_5_8 <-
  llply(
    RawNovorMix_5_8,
    filterNovor,
    ppmErrorMin = -15,
    ppmErrorMax = 15,
    minLength = 6,
    minNovorScore = 70
  )
```

# Digesting peptides and writing output files
`makePepFile` calls `digestPeptides` to make a unique list of peptides with no
missed cleavages, then writes these files to a csv file in the outputFileDir
directory. It returns a dataframe with a single row which contains the name of
the output file as well as a count of how many peptides it contains.

Here `ldply` calls `makePepFile` on each entry in `filtNovorMix_5_8` and binds
the results into a dataframe. This dataframe is merged back into
multidataTracker to add how many digested peptides resulted from each simulation
and the name of the corrisponding output file.
```{r eval=FALSE}
digestPepTable <-
  ldply(
    filtNovorMix_5_8,
    makePepFile,
    datasetCol = "SimDataset",
    sequenceCol = "cleanseq",
    outputFileDir = "./TestFolder",
    .id = "SimID"
  )

multidataTracker <- merge(multidataTracker, digestPepTable, by="SimID")
```

You can inspect resulting dataframe like so:
```{r eval=FALSE}
View(MakeSearchSim::multiDataTracker_SimID5_8)
```

