---
title: "Example workflow using simulated mzMLs with multiple contributor files with specific proteins"
author: "Isabelle OBryon"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: TRUE
vignette: >
  %\VignetteIndexEntry{Example workflow using simulated mzMLs with multiple contributor files with specific proteins}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}

---

```{r setup, include = FALSE}
library(plyr)
library(MakeSearchSim)
library(knitr)
knitr::opts_chunk$set(
  comment = "#"
)
```


# Overview

This vignette covers how to create a new mzML file that combines contains scans
for specific proteins, defined by the user, with a primary mzML file. Details of
the functions used in this process can be found in 
[MakeSearchSim Package Overview](makesearchsim-overview.html) 
to make them part of a high throughput
pipeline. It is recommended you read that vignette first.



# Inputs
1. **multi_DT**: A dataframe which contains at least the following columns. Each
row corresponds to a unique combination of primary file, contributor file, and
contributor protein.
    + Dataset: Unique name for the final mzML file, w/o the extension, and it is suggested to have the SimID in it.
    + SimID: Unique identifier. There must only be one SimID per call to MakeMultiSimFile
    + PrimaryDataset: Filename of the primary dataset w/o the extension 
    + ContributorDataset: Filename of the contributor dataset w/o the extension
    + ContributorProteins: Protein identifiers that exactly match proteins in the MS-GF+ output
    
2. **MSGF_results**:  A dataframe which contains at least the following columns.
Each row corresponds to one spectra. This dataframe should have all the PSMs for
all mzML files to be combined, and it should already be filtered if only high
quality scans from the specified proteins are wanted. The dataframe must have
the following columns:
    + Dataset: contains the dataset name
    + Protein: contains protein accessions 
    + Scan: contains scan numbers 
    + Protein: contains protein accessions 
    + **Note:** Data must be unrolled, meaning one row for each scan, peptide, and protein combination present in the data
    
3. working_dir: The location of the input mzML files and where the new mzML
files will be created.


# Example input

```{r eval=FALSE}
contrib_datasets <- c("Sample_1",
                      "Sample_2",
                      "Sample_3")

contrib_proteins <- c("FEL1B_FELCA", 
                      "TRPV6_HUMAN",
                      "CTR2_HUMAN")

primary_dataset <- c("Primary_sample",
                     "Primary_sample",
                     "Primary_sample")

example_multi_taxa_DT <-
  data.frame(
    "Dataset" = c("SimMulti_1", "SimMulti_1", "SimMulti_1"),
    "SimID" = c(1, 1, 1),
    "PrimaryDataset" = primary_dataset,
    "ContributorDataset" = contrib_datasets,
    "ContributorProteins" = contrib_proteins,
    stringsAsFactors = F
  )
mzML_dir <- "C:/Users/XXX123/my_mzML_dir"

```

# Example running
If the users want to run MakeMultiSimFile to create multiple simulated files,
they must split the input dataframe (multi_DT) by SimID and call
MakeMultiSimFile on a dataframe that only contains the information for one
simulated file.

```{r eval = FALSE}

multi_taxa_DT_finished <- plyr:ddply(example_multi_taxa_DT, 
                                     .(SimID), 
                                     MakeMultiSimFile,
                                     test_set_MSGF,
                                     mzML_dir)

```

# Output
1. A dataframe which contains at least the following columns. Each row
corresponds to a unique combination of primary file, contributor file, and
contributor protein.
    + Dataset: Unique name for the final mzML file, w/o the extension
    + SimID: Unique identifier  
    + PrimaryDataset: Filename of the primary dataset w/o the extension 
    + ContributorDataset: Filename of the contributor dataset w/o the extension
    + ContributorProteins: Protein identifiers that exactly match proteins in the MS-GF+ output
    + Scans: This column will be a 'list' datatype and it will contain all of the scans from the specified
    contributor that are in the new mzML file. 
    + SimFileName: The name of the new mzML file
    
