---
title: "Simulation workflow using simulated mzMLs"
author: "Sarah C. Jenson"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: TRUE
vignette: >
  %\VignetteIndexEntry{Simulation workflow using simulated mzMLs}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
library(stringi)
library(MakeSearchSim)
library(knitr)
knitr::opts_chunk$set(
  comment = "#"
)
```

# Overview

This vignette covers the use of the various wrapper functions which build on the
basic functions described in 
[MakeSearchSim Package Overview](makesearchsim-overview.html) 
to make them part of a high throughput pipeline. It is recommended you read that
vignette first.

This vignette describes a workflow for making simulated peptide lists by
creating a new mzML file from a primary mzML file and scans from a contributor
mzML file. Novor is then run on the new file and the results are used to make
the peptide lists. This workflow is quite slow because of the time required to
write a new mzML and run it though Novor. Running Novor will also take up more
than 90% of your CPU. If new mzMLs are not required for additional analysis, use
the workflow described in 
[Simulation workflow using existing Novor results](simulate-via-existing-novor.html) 
which is much faster because it simply mixed pre-existing Novor PSMs.

You can browse all the available vignettes by running:
```{r eval=FALSE}
browseVignettes("MakeSearchSim")
```

# Inputs
1. **A csv file(SimRecipes)** which contains at least the following columns.
Each row corresponds to a simulated dataset.
    + SimID: Unique identifier  
    + PrimaryDataset: Filename of the primary dataset w/o the extension 
    + ContributorDataset: Filename of the contributor dataset w/o the extension
    + PrimaryOrganism: name of the organism associated with the primary dataset
    + ContributorOrganism: name of the organism associated with the contributor dataset
    + TargetNumContributorProteins: How many proteins to add from the contributor dataset  
    + TargetContributorPep_per_Prot: How many peptides per protein to add from each selected contributor protein  

2. **MSGF+ results** as a list of dataframe or a single dataframe with a dataset
column that can be converted to a list. Each entry in the list contains the
MSGF+ results for a contributor dataset. The name of the dataframe in the list
must match to the entry in the ContributorDataset column of SimRecipes. The
dataframe must have the following columns:
    + Dataset: contains the dataset name
    + cleanseq: contains a clean peptide sequence (no modification info)
    + ScanNum: contains scan numbers 
    + Protein: contains protein accessions 
    + **Note:** Data must be unrolled, meaning one row for each scan, peptide, and protein combination present in the data
    
3. **mzML files** for the primary and contributor datasets. 

**Inputs for this vignette:**
The mzML files are very big and the pipeline takes a while to run, so for this
vignette I've included an input SimRecipes table(SimRecipes_SimID5_8), and the
dataTracker object you would get at the end(dataTracker_Sim5_8) which I'll
reference throughout.

# Creating Simulated Datasets

#### Load Inputs
If the contributor MSGF+ results are already split into a list of dataframes
then you're good to go. However if they're in one big dataframe you can make a
list of dataframes by spliting them by a column containing the dataset name like
so:

```
dbResultList <- split(alldbresults, alldbresults$Dataset)
```
Here's an example `SimRecipes` table included with the package.
```{r Load-SimRecipes, echo=TRUE}
SimRecipes <- MakeSearchSim::SimRecipes_SimID5_8
```


You'll need to have a column called 'SimFileName' in the SimRecipes dataframe
specifying the filename of the simulated file you're going to create.

The filename is meant to  be a user readable summary of what's in the file. Here
I'm using the first 6 characters of the primary and contributor organisms
folowed by the number of contributor proteins and peptides/protein. Then I add
the SimID at the end.

You can use whatever format you like as long as the SimID is clearly denoted
somewhere in the filename.

```{r Generate-Simulated-File-Name, eval=FALSE}
#adding Simulated File name
#(first 6 char of primary)_(first 6 char of contributor)_Nprot(#contributor proteins)
#_Npep(Number of peptides per protein)_SimID(SimID number).mzML
#Ex: Pseudo_Alcali_Nprot10_Npep5_SimID1.mzML

SimRecipes$SimFileName <-
  paste0(
    substr(SimRecipes$PrimaryOrganism, 1, 6),
    '_',
    substr(SimRecipes$ContributorOrganism, 1, 6),
    '_Nprot',
    SimRecipes$TargetNumContributorProteins,
    '_Npep',
    SimRecipes$TargetContributorPep_per_Prot,
    '_SimID',
    SimRecipes$SimID,
    ".mzML"
  )

knitr::kable(SimRecipes)
```


#### Running MakeSimFile()
`MakeSimFile()` is basically a wrapper function for lower level functions
`selectScans()` and `SubsetMergeMzmls()` with some input checking and assertions
added. It is meant to be used with `plyr::ddply()` to iterate over each row of
`SimRecipes`. The output is the row of `SimRecipes` with added columns
containing what contributor scans, peptides, and proteins were added along with
the date and time the file was made. `plyr::ddply()` combines all the new rows into a
dataframe called `dataTracker`.

Contributor scans are selected according to the value of
`TargetContributorPep_per_Prot` and `TargetNumContributorProteins.` The
`sample()` function is used to randomly select a certain number of proteins and
then a certain number of peptides from those proteins. The SimID is used as the
seed for `sample()` to ensure reproducibility. So you will get the exact same
simulated mzML each time you run the code given the same `SimRecipes` row.

__Note:__ It is possible to end up with more contributor proteins than the
`SimRecipes` specifies if a selected peptide belongs to more than one protein.

```{r MakeSimFiles, eval=FALSE}
dataTracker <- ddply(SimRecipes, 
                   .(SimID), 
                   MakeSimFile, 
                   dbResults = dbResultList, 
                   rng_version = "R_version_default")
```

`dataTracker` now contains new columns detailing what scans, peptides, and
proteins were taken from the contributor to make the simulated dataset.

```{r}
colnames(MakeSearchSim::dataTracker_SimID5_8_mzMLex)[10:16]
```

The `ContributorScans`, `ContributorPeptides`, and `ContributorProteins` columns are
technically lists of vectors so you can extract the values like so:

```{r}
#extracting contributor proteins within SimID 5
MakeSearchSim::dataTracker_SimID5_8_mzMLex$ContributorProteins[1]
```

# Running Novor
`runNovorSingle()` is meant to be run with `ddply()` over each row in `dataTracker`.
It first calls `mzMLtoMGF()` to convert the simulated mzML to an MGF file using
MSConvert. Then it calls `runNovor()` to run Novor on the new MGF file via the
command line.

Both MSConvert and Novor require configuration files which are passed to
`runNovorSingle()` as additional arguments in the `ddply()` call. Another
additional argument is the path to the local Novor `.bat` file. It must be a
windows style path to the Novor `.bat` file within the Novor installation
folder. Backslashes must be escaped by a second backslash. Ex:
`C:\\Novor\\win\\novor.bat`. You can generate an appropriate path using the
`choose.files()` function which will allow you to use a GUI.

```{r Getting-config-files, eval=FALSE}
msConvert_config <- system.file("extdata", 
                                "convertMGF_config.txt", 
                                package = "MakeSearchSim", 
                                mustWork = TRUE)

novor_config <- system.file("extdata",
                            "NovorParams_fTol002_pTol15_MetOx_noAlk_Novor_params.txt",
                            package = "MakeSearchSim",
                            mustWork = TRUE)

path_to_novor <- choose.files()
```

`runNovorSingle()` calls `mzMLtoMGF()` which requires an output directory for
the MGF files to be specified by using the `mgf_output_dir` parameter.
```{r Running-Novor, eval=FALSE}
#running runNovorSingle() on each row of dataTracker
dataTracker<-ddply(dataTracker, 
                   .(SimID), 
                   runNovorSingle,  
                   msConvert_config = msConvert_config, 
                   novor_config = novor_config, 
                   Novor_Path = path_to_novor, 
                   mgf_output_dir = ".")
```

`runNovorSingle()` returns an updated `dataTracker` row which contains extra
columns for the MGF file name, raw Novor results file, and the Novor parameter
file name. `ddply()` binds these rows together to make a new `dataTracker`.

These are the names of the new columns.
```{r}
colnames(MakeSearchSim::dataTracker_SimID5_8_mzMLex)[17:19]
```

## Importing and Filtering Novor Results

### importFilter_Novor()  

Novor results will be in a csv whose name is the base name of the simulated file
with `_Novor` appended. This filename is stored in `dataTracker` under the
`Raw_Novor_Filename` column.

The `importFilter_Novor()` function is meant to be called with `plyr::dlply()`
on the `dataTracker` dataframe. It calls two other functions,
`importSingleNovor()` and `filterNovor()`, to add metadata and filter the Novor
results.

It returns a list of dataframes, each dataframe corresponding to a row in
`dataTracker`. See 
[MakeSearchSim Package Overview](makesearchsim-overview.html) for more details
about how Novor results are imported. 

Below is an example call using the default filtering parameters of ppm +/- 15,
minimum Novor score 70, and minimum length 6.
```{r importFilter_Novor, eval=FALSE}
SimResultList <- dlply(dataTracker, .(SimID), importFilter_Novor)
```

You can specify different filtering parameters like so:
```{r alt-importFilter_Novor, eval=FALSE}
SimResultList <-dlply(dataTracker, 
                      .(SimID), 
                      importFilter_Novor, 
                      ppmErrorMin = -10, 
                      ppmErrorMax = 10, 
                      minNovorScore = 80, 
                      minLength = 7)
```

__Note:__ These will be used for every file represented in dataTracker.

# Digesting Peptides and writing output files
`makePepFile()` takes a dataframe containing at least a column of peptide
sequences and a column containing the dataset name and writes a `.csv` file
containing the unique list of peptides resulting from digesting the peptides
with `digestPeptides()`. See the 
[MakeSearchSim Package Overview](makesearchsim-overview.html) vignette 
or the help page for a description of `digestPeptides()`.

The output csv file contains a single column of peptides with `x` as the first
entry. The filename is the first entry in the `datasetCol` column with
`_peptides.csv` added. The function returns a dataframe row containing the name
of the output file along with the number of peptides contained in the file. You
can also select an output directory by supplying a relative file path from the
working directory.

You can call `makePepFile()` on each entry in `SimResultList` using `ldply()`
which will result in table which you can then merge with `dataTracker`. You can
use the `outputFileDir` parameter with a file path relative to the current
working directory to set the folder where the files will be written.
```{r eval=FALSE}
digest_pep_table <- ldply(SimResultList, 
                          makePepFile, 
                          prolineBlock = TRUE, 
                          discardLessThan = 6, 
                          datasetCol = "MGFDataset")

#merging back into dataTracker to add number of digested peptides
dataTracker <- merge(dataTracker, digest_pep_table, by = "SimID", all = TRUE)
```

Here's the added columns showing how many digested peptides we got from each file.

```{r}
knitr::kable(MakeSearchSim::dataTracker_SimID5_8_mzMLex[,20:21])
```

When you're done save `dataTracker` as an `Rdata` object with a descriptive variable
name and filename.

```{r eval=FALSE}
dataTracker_Sim5_8<-dataTracker
save(dataTracker_Sim5_8, file="dataTracker_Sim5_8.RData")
```


