---
title: "Simulation workflow using existing Novor results"
author: "Sarah C. Jenson"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: TRUE
vignette: >
  %\VignetteIndexEntry{Simulation workflow using existing Novor results}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
library(plyr)
library(reshape2)
library(assertthat)
library(knitr)
library(MakeSearchSim)

knitr::opts_chunk$set(
  comment = "#"
)
```

# Overview
This vignette describes a workflow for making simulated peptide lists by mixing
Novor results from a contributor and primary dataset. This workflow bypasses the
creation of new mzMLs and the need to run Novor on each one. This results in a
much faster workflow. 
See [Simulation workflow using simulated mzMLs](simulate-via-mzml.html) 
for a workflow which creates new mzMLs.

This vignette covers the use of the various functions which build on the basic
functions described in 
[MakeSearchSim Package Overview](makesearchsim-overview.html) to make them part 
of a high throughput pipeline. It is recommended you read that vignette first.


# Inputs
1. **A csv file(SimRecipes)** which contains at least the following columns.
Each row corresponds to a simulated dataset.
    + SimID: Unique identifier  
    + PrimaryDataset: Filename of the primary dataset w/o the extension 
    + ContributorDataset: Filename of the contributor dataset w/o the extension
    + PrimaryOrganism: name of the organism associated with the primary dataset
    + ContributorOrganism: name of the organism associated with the contributor dataset
    + TargetNumContributorProteins: How many proteins to add from the contributor dataset  
    + TargetContributorPep_per_Prot: How many peptides per protein to add from each selected contributor protein  

2. **MSGF+ results** as a list of dataframe or a single dataframe with a dataset
column that can be converted to a list. Each entry in the list contains the
MSGF+ results for a contributor dataset. The name of the dataframe in the list
must match to the entry in the ContributorDataset column of SimRecipes. The
dataframe must have the following columns:
    + Dataset: contains the dataset name
    + cleanseq: contains a clean peptide sequence (no modification info)
    + ScanNum: contains scan numbers 
    + Protein: contains protein accessions 
    + **Note:** Data must be unrolled, meaning one row for each scan, peptide, and protein combination present in the data
    
3. **Novor results:** This vignette describes how to run Novor using functions
included in MakeSearchSim, however you can also use Novor results you already
have and skip skip to the [Importing Novor Results](#importnovor) section.

# Generating Novor results

#### mzML to MGF conversion
If you do not already have MGF files, you can convert mzML files to MGF using
`mzMLtoMGF()` which calls MSConvert from the command line. If you have make the
MGF files elsewhere, they must have titles in the same format as created by
MSConvert using this command:

```
filter="titleMaker <RunId>.<ScanNumber>.<ScanNumber>.<ChargeState> File:"<SourcePath>", NativeID:"<Id>""
```

```{r eval=FALSE}
mzMLFiles <- list.files(pattern = "*.mzML", full.names = TRUE)

msConvert_config <-
  system.file("extdata",
              "convertMGF_config.txt",
              package = "MakeSearchSim",
              mustWork = TRUE)

mgfFiles <-
  lapply(mzMLFiles, mzMLtoMGF, configFile = msConvert_config)
```


## Running Novor
You can run Novor using `runNovor()`, included in the package are parameter
files appropriate for high/high resolution QExactive data. Below uses a Novor
parameter file for non-alkylated samples (no fixed cysteine carbamidomethylation
modification). A parameter for for alkylated samples (with a fixed cysteine
carbamidomethylation modification) called
NovorParams_fTol002_pTol15_MetOx_Alk_Novor_params.txt is also included with the
package and can be accessed in the same way. Both parameter files contain a
variable methionine oxidation modification.

Another additional argument is the path to the local Novor .bat file. It must be
a windows style path to the Novor .bat file within the Novor installaion folder.
Backslashes must be escaped by a second backslash. Ex:
`C:\\Novor\\win\\novor.bat`.

You can generate an appropiate path using the `choose.files()` function which
will allow you to use a GUI.

```{r eval=FALSE}
novor_config <-
  system.file(
    "extdata",
    "NovorParams_fTol002_pTol15_MetOx_noAlk_Novor_params.txt",
    package = "MakeSearchSim",
    mustWork = TRUE
  )

path_to_novor <- choose.files()

novorResFiles <-
  sapply(mgfFiles,
         runNovor,
         NovorParams = novor_config,
         Novor_Path = path_to_novor)
```

## Importing Novor Results {#importnovor}
`importSingleNovor()` reads in Novor's output csvs, cleans up column names, then
calls `SingleMGFIDtoScan()` to add scan numbers. See the associated help pages
for more details. Using `mdply` imports all the Novor results as a single
dataframe. You can then split that dataframe into a list by a column containing
the dataset name. This makes each entry in the have the name of the dataset it
came from, which is necessary later on.
```{r eval=FALSE}
novorResFiles <- list.files(pattern = "*._Novor.csv")
mgfFiles <- list.files(pattern = "*.mgf")

#making a dataframe of result csv's and corrispoinding MGF files
inputDF <-
  cbind.data.frame(
    NovorOutputFile = novorResFiles,
    MGF_file = unlist(mgfFiles),
    stringsAsFactors = FALSE
  )

#running importSingleNovor
RawNovor <- mdply(inputDF, importSingleNovor)
RawNovor$MGFDataset <- factor(RawNovor$MGFDataset)

#spliting into a list of dataframes by MGFDataset
RawNovorList <- split(RawNovor, RawNovor$MGFDataset)
```

# Selecting contributor scans
First load the MSGF+ results for the contributor files. This must be in the form
of a list, with each entry named after it's dataset and containing a dataframe.

Loading MSGF+ results and SimRecipes included in package
```{r eval=FALSE}
load("dbResultList_Sim5_8.RData")
SimRecipes <- MakeSearchSim::SimRecipes_SimID5_8
```

Adding column for dataset name: This workflow requires a column for the desired
simulated dataset name. This is used to name the output peptide lists. This is
different than the SimFileName used in the 
[simulated mzML workflow](simulate-via-mzml.html) in that it does not contain
the ".mzML" extension.
```{r eval=FALSE}
SimRecipes$SimDataset <-
  paste0(
    substr(SimRecipes$PrimaryOrganism, 1, 6),
    '_',
    substr(SimRecipes$ContributorOrganism, 1, 6),
    '_Nprot',
    SimRecipes$TargetNumContributorProteins,
    '_Npep',
    SimRecipes$TargetContributorPep_per_Prot,
    '_SimID',
    SimRecipes$SimID
  )
```

## Running BatchSelectScans
`BatchSelectScans()` is a wrapper function for `selectScans()` and is meant to
be used with `ddply()` on a SimRecipes table and a list of contributor MSGF+
results such that `BatchSelectScans()` is called once for each row in the
SimRecipes table. First, `BatchSelectScans()` translates the parameters stored
in a row of the `SimRecipes` table to inputs for `selectScans()`. It then uses
the scan numbers returned by `selectScans()` to collect information about their
corresponding peptides and proteins from the contributor MSGF+ results. It then
creates a one row dataframe containing this data as well as whatever columns
were in the input row of `SimRecipes.` `ddply()` binds the results from each
iteration of `BatchSelectScans()` to create a `dataTracker` dataframe.

The structure of the `dataTracker` dataframe is covered in more detail in
[Simulation workflow using simulated mzMLs](simulate-via-mzml.html). 

```{r eval=FALSE}
dataTracker <-
  ddply(
    SimRecipes,
    .(SimID),
    BatchSelectScans,
    dbResults = dbResultList_Sim5_8,
    rng_version = "R_version_default"
  )
```


# Mixing primary and contributor Novor results

There are two workflows for mixing primary and contributor Novor results, the
small set workflow is for sets of about 50 or fewer simulations. See the 
[large set workflow](#largeset) for large sets of simulations.

## Small set workflow
This workflow uses `BatchMixNovor()`, `filterNovor()`, and `makePepFile()` one
at a time by holding the results of each step in memory using a list of
dataframes. This allows you to have the mixed and filtered Novor results
available in the global environment for further inspection. However this takes
up a lot of memory for large sets of simulations.

`BatchMixNovor()` is a wrapper of `MixNovor()` which takes parameters for
`MixNovor()` from `dataTracker_row`. This is necessary so you can call `ddply()`
on a `dataTracker` dataframe to make batches of simulated Novor results. After
calling `MixNovor()` it adds a column defining the simulated dataset name which
will be needed to make peptide list files. It gets this value from the
`SimDataset` column in `dataTracker_row.`

`MixNovor()` extracts the Novor results of the scan numbers in
`contributorScans` and adds them to the Novor results of the primary. **Note:**
To avoid duplication of scan numbers, it is recommended to have a column in the
input Novor result dataframes which lists the original file. This is added if
you use `importSingleNovor()` to import the Novor result from the csv files.

```{r eval=FALSE}
allmixedNovor <- ddply(dataTracker, 
                       .(SimID), 
                       BatchMixNovor, 
                       NovorResultList = RawNovorList)
```

**Note:** Just because a scan has a high quality PSM in the MSGF+ results,
doesn't mean Novor will return a PSM for that scan. This means Novor may not
return any PSMs for some of the contributor scans.
 

### Filtering Novor results
`filterNovor()` is a helper function which filters raw Novor results by user
specified cutoffs of ppm error, Novor Score, and peptide length. Peptide length
is calculated within the function based on the `cleanseq` column and added as a
`length` column.

Below shows a call with the default values.
```{r eval=FALSE}
filtmixedNovor <- filterNovor(allmixedNovor,
                              ppmErrorMin = -15,
                              ppmErrorMax = 15,
                              minLength = 6, 
                              minNovorScore = 70)
```

### Writing Peptide list files
`makePepFile()` takes a dataframe containing at least a column of peptide
sequences and a column containing the dataset name and writes a .csv file
containing the unique list of peptides resulting from digesting the peptides
with `digestPeptides()`. 
See the [MakeSearchSim Package Overview](makesearchsim-overview.html) vignette
or the help page for a description of `digestPeptides()`.

The output csv file contains a single column of peptides with `x` as the first
entry. The filename is the first entry in the `datasetCol` column with
`_peptides.csv` added. The function returns a dataframe row containing the name
of the output file along with the number of peptides contained in the file. You
can also select an output directory by supplying a relative file path from the
working directory.


```{r eval=FALSE}
digestPepCounts <-
  ddply(filtmixedNovor, 
        .(SimID), 
        makePepFile, 
        datasetCol = "SimDataset")

#Merging back with dataTracker
dataTracker <- merge(dataTracker, digestPepCounts, by = "SimID")
```

## Large set workflow {#largeset}
This workflow is for large sets of simulations. It uses `ddply` with
`MixFilterDigest()`, which is a wrapper function that calls `MixNovor()`,
`filterNovor()`, and `makePepFile()` given a single `dataTracker` style row.
This means that at any given time only the Novor results for one simulation are
held in memory. The downside of using this function is that you can't
interrogate the the raw and filtered mixed Novor results for each simulation.

`MixFilterDigest()` returns a `dataTracker` row with additional columns
containing the name of the peptide list file (`PeptideListFile`) and the number
of digested peptides (`Num_Digested_Peptides`).

```{r eval=FALSE}
dataTracker <-
  ddply(
    dataTracker,
    .(SimID),
    MixFilterDigest,
    NovorResultList = RawNovorList,
    outputFileDir = "./PepListFolder"
  )
```





